---
title: "ICD-10-PCS"
format: html
---

The __ICD-10-PCS__ (International Classification of Diseases, 10th Revision, Procedure Coding System) is a procedure classification published by the United States for classifying procedures performed in hospital inpatient health care settings. The procedure codes have been adopted under HIPAA for hospital inpatient healthcare settings.

```{r}
#| message: false
#| warning: false
library(icd10us)
library(tidyverse)
library(gt)
library(janitor)
```

# Code Conventions

ICD-10-PCS codes are composed of seven characters. Each character is an axis of classification that specifies information about the procedure performed. Within a defined code range, a character specifies the same type of information in that axis of classification.
Example: The fifth axis of classification specifies the approach in sections 0 through 4 and 7 through 9 of the system.

ICD-10-PCS has a __7-character__ alphanumeric code structure. Each character contains up to __34__ possible values. The seven characters that make up a complete code have specific meanings that vary for each of the __17__ sections of the manual.


```{r}
format(7^34, scientific = FALSE, big.mark = ",") # possible combinations
```


Each value represents a specific option for the general character definition. The 10 digits `0:9` and the 24 letters A-H, J-N, and P-Z may be used in each character.

```{r}
#| message: false
#| warning: false

# "[0-9A-HJ-NP-Z]" regex pattern

valid_chars <- c(0:9,
                 LETTERS[1:8],     # A-H
                 LETTERS[10:14],   # J-N
                 LETTERS[16:26])   # P-Z

# length(valid_chars)

stringr::str_flatten_comma(valid_chars, last = ", and ")
```


The letters `O` and `I` are not used so as to avoid confusion with the digits `0` and `1`.


```{r}
#| message: false
#| warning: false
pcs.structure <- tibble(
  Character = 1:7,
  Meaning = c('Section',
              'Body System',
              'Root Operation',
              'Body Part',
              'Approach',
              'Device',
              'Qualifier')) |> 
  datawizard::data_rotate()

pcs.structure[2, ] |> 
  gt() |> 
  cols_align(align = 'center') |> 
  opt_table_font(font = google_font(name = "Fira Code")) |> 
  opt_table_lines(extent = "all") |> 
  gtExtras::gt_highlight_cols(1, 
                              fill = 'yellow', 
                              alpha = 0.5, 
                              font_weight = 'bold') |> 
  gtExtras::gt_add_divider('1', sides = 'all', style = 'solid')
```

# Section

```{r}
spl <- icd10pcs |> 
  filter(valid_billing_code == TRUE) |> 
  select(icd10pcs_code, icd10pcs_long_description) |> 
  pull(icd10pcs_code) |> 
  str_split_fixed("", n = 7) |> 
  as.data.frame()

spl <- dplyr::tibble(
  section = spl$V1,
  body_system = spl$V2,
  root_operation = spl$V3,
  body_part = spl$V4,
  approach = spl$V5,
  device = spl$V6,
  qualifier = spl$V7
)
```


```{r}
tabyl(spl, section) |> 
  untabyl() |> 
  left_join(
    tibble(section = icd10pcs_sections$section_digit,
           description = icd10pcs_sections$section_description), 
    by = join_by(section)) |> 
  select(section, description, count = n, percent) |> 
  arrange(desc(count)) |> 
  gt(rowname_col = "section") |> 
  fmt_percent(columns = percent) |> 
  cols_align(align = 'center', columns = section) |> 
  opt_table_font(font = google_font(name = "Fira Code")) |> 
  opt_table_lines(extent = "all") |> 
  opt_align_table_header("left") |> 
  tab_header(title = "ICD-10-PCS 2024: Breakdown by Sections")
```



## Medical and Surgical (`0`)

```{r}
#| message: false
#| warning: false
tibble(
  Meaning = c('Section',
              'Body System',
              'Root Operation',
              'Body Part',
              'Approach',
              'Device',
              'Qualifier'),
  # Character = 1:7,
  Code = c("0", rep("x", 6))) |> 
  datawizard::data_rotate() |> 
  gt() |> 
  cols_align(align = 'center') |> 
  opt_table_font(font = google_font(name = "Fira Code")) |> 
  opt_table_lines(extent = "all") |> 
  gtExtras::gt_highlight_cols(1, 
                              fill = 'yellow', 
                              alpha = 0.5, 
                              font_weight = 'bold') |> 
  gtExtras::gt_add_divider('1', sides = 'all', style = 'solid')
```

## Body System

```{r}
filter(spl, section == "0") |> 
  tabyl(body_system) |> 
  untabyl() |> 
  mutate(description = map_chr(body_system, system.medsurg)) |> 
  select(body_system, 
         description, 
         count = n, 
         percent) |> 
  arrange(desc(count)) |> 
  gt(rowname_col = "body_system") |> 
  fmt_percent(columns = percent) |> 
  fmt_integer(columns = count) |> 
  cols_align(align = 'center', columns = body_system) |> 
  opt_table_font(font = google_font(name = "Fira Code")) |> 
  opt_table_lines(extent = "all") |> 
  opt_align_table_header("left") |> 
  tab_header(title = md("**ICD-10-PCS 2024**: Body Systems"),
             subtitle = md("Section **[0]** Medical & Surgical"))
```

## Root Operation

```{r}
tabyl(spl, root_operation, section) |> 
  adorn_totals() |>
  adorn_totals("col") |>
  #adorn_percentages() |> 
  #adorn_pct_formatting() |> 
  #adorn_ns() |> 
  adorn_title()
```

<br>

```{r}
filter(spl, section == "0") |> 
  tabyl(body_system, root_operation) |> 
  untabyl() |> 
  mutate(description = map_chr(body_system, system.0)) |> 
  select(system = description,
         bs = body_system,
         "Alteration [0]" = `0`,
         "Bypass [1]" = `1`,
         "Change [2]" = `2`,
         "Control [3]" = `3`,
         "Creation [4]" = `4`,
         "Destruction [5]" = `5`,
         "Detachment [6]" = `6`,
         "Dilation [7]" = `7`,
         "Division [8]" = `8`,
         "Drainage [9]" = `9`,
         "Excision [B]" = `B`,
         "Extirpation [C]" = `C`,
         "Extraction [D]" = `D`,
         "Fragmentation [F]" = `F`,
         "Fusion [G]" = `G`,
         "Insertion [H]" = `H`,
         "Inspection [J]" = `J`,
         "Map [K]" = `K`,
         "Occlusion [L]" = `L`,
         "Reattachment [M]" = `M`,
         "Release [N]" = `N`,
         "Removal [P]" = `P`,
         "Repair [Q]" = `Q`,
         "Replacement [R]" = `R`,
         "Reposition [S]" = `S`,
         "Resection [T]" = `T`,
         "Supplement [U]" = `U`,
         "Restriction [V]" = `V`,
         "Revision [W]" = `W`,
         "Transfer [X]" = `X`,
         "Transplantation [Y]" = `Y`) |> 
  gt() |> 
  cols_align(align = 'right', columns = bs) |>
  cols_align(align = 'center', columns = !bs) |>
  cols_label(bs = "System") |> 
  opt_table_font(font = google_font(name = "Fira Code")) |> 
  opt_table_lines(extent = "all") |> 
  opt_align_table_header("left") |> 
  tab_header(title = md("**ICD-10-PCS 2024**: Root Operations"),
             subtitle = md("Section **[0]** Medical & Surgical by **Body Systems**")) |> 
  sub_zero(zero_text = ".") |> 
  gtExtras::gt_merge_stack(bs, system) |> 
  opt_all_caps()
```

## Approach

```{r}
filter(spl, section == "0") |> 
  tabyl(root_operation, approach) |> 
  untabyl() |> 
  arrange(desc(n))
```

